### 2012-11-07 thesis/tex/6_theltlcreatorprototype.tex
Almost all correction issues of Andreas and Alwin are realized now (except the ones of Alwin I couldn't read). Chapter 4 and 6 have a new structure.

First of all some details about the implementation of the visual language are given in section~\ref{sec:prototype:operatorconstraints}, directly followed by an explanation about the realization of the automated constraint generation functionality in section~\ref{sec:prototype:automatedconstraintgeneration}.
The assembly of all the previous features in one tool is described in section~\ref{sec:coalescence}. However, the tool's functionality shall also be available in Robostudio in order to provide safety mechanisms for the healthcare service robots and to be a prototype for evaluation. Section~\ref{sec:integrationintorobostudio} gives a main idea how an integration into Robostudio is possible.

@@

\begin{description}
	\item[getLTL():] This function returns the LTL formula of an operator and all its children recursively. Applied to the most outer operator it returns the complete formula for the constraint. This function is abstract and implemented by each operator type separately.
	\item[getColor():] Each operator type has got is own significant color which can be retrieved by the abstract function \emph{getColor()}.
	\item[createNewInstance():] In order to provide operator creation over a tool bar, this function instantiates a new operator object, depending on the implementation.
	\item[setMouseOver(boolean):] Mouse over actions cause the underlying operator to be highlighted. This function makes all operators except the hovered one appearing bleeched out.
	\item[paintOperator(Graphics, ...):] Each operator class has to implement the abstract function \emph{paintOperator()}. It manages all color and shape painting for the respective operator. Additional parameters such as recursive painting including children can be specified.
	\item[contains(int, int):] Given two coordinates x and y, it can be determined whether the herewith defined point lays within an operator.
	\item[isSimilar(AbstractOperator):] Two operators can be compared for semantic equality. Equality is given when the compared operators are of the same type and similarly their children. This functionality is important for avoiding duplicate constraints during constraint generation.
	\item[addChangeListener(OperatorChangeListener):] Whenever there is a change to the operator or one of its children, an operator change listener gets notified about it. The hierarchical architecture of \emph{AbstractOperator}s propagates all change events to the root operator. The function \emph{addChangeListener()} allows registering for such change notifications.
	\item[removeChangeListener(OperatorChangeListener):] According to \emph{addChangeListener} this function allows deregistration again.
\end{description}

@@

Whereas \emph{StateOperator}s have got no operator children and are leafs in the syntax tree of LTL formulas, all other operator types are either unary or binary operators with one or two possible children respectively. 
The graphical representation of both unary and binary operators thus needs to have place holders and docking stations for sub operators. These are realized by buckets which display a small ``drop here'' label and allow operator adding or removing by mouse drops. For this a class Bucket is defined with get and set methods for defining a sub operator.
Figure~\ref{fig:klassendiagramm_operators} depicts the relationship between \emph{AbstractOperator}, \emph{Bucket} and operator implementations.
\begin{figure}[htbp]
  \centering
  \includegraphics[width=\linewidth]{klassendiagramm_operators} 
  \caption{Class diagram of operator structure.}
  \label{fig:klassendiagramm_operators}
\end{figure}
Once operators are composed in a way that no empty buckets are left over, we retrieve a complete constraint such as the one shown in figure~\ref{fig:operator_tree}. This diagram illustrates the correlation of nested operators and buckets.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\linewidth]{operator_tree} 
  \caption{Instantiation of a constraint with several operators.}
  \label{fig:operator_tree}
\end{figure}

All graphical components such as buckets and operators are implemented as Java Swing \emph{JComponent}s. Thereby a lot of functionality for the visual configuration is already available such as the complex layout and repaint concept and the container functionality which is used for unary and binary operators as well as buckets.

@@

For lightweight components Swing allows the implementation of a \emph{paint()} method which is responsible for rendering the component. This is used for giving every operator its special look. We hereby made use of Swing's \emph{Graphics2D} functionality what gives a great support in painting rounded shapes and color gradients. An operator is displayed as a rounded rectangle with dents around sub operators and has a flowing backgound, borders and labels in an operator type specific color. For the significant shape multiple rounded rectangles with and without border are arranged in a way that the result appears to be one piece. Figure~\ref{fig:paintprocess} demonstrates how the process of painting is: For painting the \emph{IfThenOperator}, first of all a blank rectangle with border is drawn, directly followed by two filled rectangles with border which are placed where later the two child operators shall be. In the third step the same rectangle as from step one is painted on top, this time without border. This makes all crossing borders disappear. Only labels and buckets still have to be added to create a complete \emph{IfThenOperator}.

@@

Swing also provides a mechanism for layouting components, that means determining the aspired sizes and positions of components depending on child components. As known from figure~\ref{fig:operators} in section~\ref{sec:operatorconstraintformalism}, constraints have particular meanings for the two dimensions: a vertical read direction and the horizontal direction for variation in time. This fact forms a special requirement for the layouting of operators.
Whereas the operator's positioning and resizing along the logical direction only depends on the sizes of sub operators, the arrangement within the time axis turns out to be more difficult. Here sub operators can't just be aligned centered as it is the case vertically. To face the determination of horizontal positioning, every operator gets time reference lines as depicted in figure~\ref{fig:referencelines}. Logical operators, i.e. \emph{IfThenOperator}, \emph{AndOperator}, \emph{OrOperator}, \emph{NotOperator} and \emph{StateOperator}, have one reference line, time relevant operators, i.e. \emph{NextOperator}, \emph{FutureOperator}, \emph{AlwaysOperator} and \emph{UntilOperator}, have two such reference lines.

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.65]{referencelines} 
  \caption{Time reference lines and their adjustments.}
  \label{fig:referencelines}
\end{figure}

@@

The built-in mouse support of Swing was used for identifying operators positioned under the mouse cursor. On the one hand this is important for context sensitive highlighting of particular constraint parts, on the other hand this way the position where operators have to be added after drag and drop can be determined. So far
% the predefined Swing drag and drop has not been used but
a simple drag and drop mechanism has been implemented whose class structure is shown in figure~\ref{fig:dndclassstructure}.
\begin{figure}[htbp]
  \centering
  \includegraphics[width=\linewidth]{dndclassstructure} 
  \caption{Class definition for drag and drop functionality.}
  \label{fig:dndclassstructure}
\end{figure}% instead. 
Once a \emph{JComponent} is registered to the \emph{DndHandler}, it automatically manages any drag and drop actions with \emph{JComponent}s which implement either the \emph{Draggable} or \emph{DropTarget} interface. The former is implemented by all operators, the latter by buckets and the dashboard.
%Figure~\ref{fig:dndclassstructure} gives a short class overview.
Furthermore there is a class \emph{DraggableProvider} which displays a kind of button in the tool bar and lets the user create new \emph{Draggable}s; in our case new operators get instantiated.

%operator werden im dashboard, einem weiteren container, zusammengeklickt und angezeigt. Das Dashboard hat dabei eine ähnliche Funktionsweise wie ein Bucket. die operatoren können dabei vom operatorprovider bezogen werden: Implementierungsdetails? trash...

@@

Section~\ref{sec:subgraphdefinition} states that automated constraint generation is based on subgraphs. For the finding of such subgraphs an algorithm has been developed. It is described as follows:

First of all one start state ($\alpha$) is defined and all possible paths are generated which start from it and lead through the state machine. Excluded are all paths which contain states more than once.
The result is a finite amount of finite paths since loops are eliminated. Now every state gets assigned a number which represents the total number of occurrences of the respective state in all collected paths. In other words, the label of a state expresses the number of paths which start in the start state and lead through the respective state. The start state itself is labelsed the number of all paths. If during walking through any of the paths a second state can be found with the same label, it means that all paths starting at start state come all together in this state again. It is determined to be a potential corresponding end state.
A last check has to be applied whether there are incoming transitions from outside into the subgraph which don't originally come from start state. If no such transitions are detected, a subgraph is found.
If there is no state with the same label, or transitions come from outside, the state machine as a whole can be assumed as subgraph beginning with start state insofar as the state machine doesn't contain further loops or terminating end states. Figure~\ref{fig:subgraph_finding} illustrates this concept of subgraph finding and shows all possible paths as well as the number of occurrence of each state.
This procedure is suitable for checking whether a state is the beginning of a subgraph and for discovering the corresponding end state. In order to detect all subgraphs within a state machine, this procedure has to be applied to every single state.

\begin{figure}[htbp]
  \centering

@@

	\end{tikzpicture}
  \caption{Example for subgraph finding. The occurrence of each state in the paths lets you determine G as an end state corresponding to A.}
  \label{fig:subgraph_finding}
\end{figure}

Following the above description an algorithm has been implemented in Java. Its realisation differs from the presented concept in one point: Paths are not listed but states are directly labelled since saving all possible paths would be memory inefficient.
The state machine is traversed similarly to a depth-first search and each visited state's label is incremented by one. If the currently visited state holds more than one outgoing transition and thus constitutes new paths,  all states on the current stack, i.e. the previously visited states on the current path, get their labels incremented by $(\textnormal{'number of outgoing transitions'} - 1)$. The algorithm turns back whenever it comes across a state which has already been visited within the same path stack of the depth-first search.

The overall implemented way of subgraph finding is illustrated in pseudo code with algorithm~\ref{alg:subgraphfinding}. In line 4 the labelling process described above is executed by using one particular start state. If a state has the same label as the start state, it is assumed as potential end state. Otherwise, the entire state machine is assumed to be a subgraph and start state is defined as end state as well. Subsequent checks finally determine whether it can be considered as proper subgraph.
There are no cycles allowed between start and end state since they're forbidden by the subgraph definition in section~\ref{sec:subgraphdefinition}.
Furthermore also foreign transitions are prohibited. Transitions coming from a state outside the subgraph are allowed to lead to the start state but not to other states of the subgraph. If all checks are successful, start and end states get added to the result list $subgraphs$. This procedure is applied to all states of the state machine by a surrounding foreach loop. In the end, the list $subgraphs$ contains all pairs of start and end states which form subgraphs.

\begin{algorithm}
\caption{Finding of all subgraphs.}
\label{alg:subgraphfinding}
\begin{algorithmic}[1]
\STATE $subgraphs$: List
\FORALL{states as $S$}
  \STATE $start \leftarrow S$
	\STATE doLabelling($start$);
	\IF{state $E$ exists with label($E$)$=$label($S$)}
		\STATE $end \leftarrow E$
	\ELSE
		\STATE $end \leftarrow start$
	\ENDIF
	\IF{noCycles($start$, $end$) $and$ noExtTransitions($start$, $end$)}
		\STATE $subgraphs$.add($start$, $end$);
	\ENDIF
\ENDFOR
\end{algorithmic}
\end{algorithm}

The runtime of the algorithm depends on size and grade of branch of the state machineh. To improve user experience during constraint generation a dialog has been implemented which informs the developer about the current progress. This includes the number of constraints found so far and a progress bar which gives a percentual estimation of execution time. Furthermore a cancel button permits premature stopping of the execution. A snapshot of the Dialog is shown in figure~\ref{fig:progressdialog}.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.40\linewidth]{progressdialog} 
  \caption{Snapshot of the progress dialog displayed during constraint generation.}
  \label{fig:progressdialog}
\end{figure}

The algorithm implementation has been extended by code snippets which frequently interact with so called \emph{ValidationResultListener}s (figure~\ref{fig:validationresultlistener_class}). This interface is also implemented by the progress dialog. A \emph{ValidationResultListener} has to provide three methods which allow the algorithm to publish the current progress by calling \emph{void setProgress(int)} or to notify about newly found subgraphs by invoking \emph{void newSubGraphFound(SubGraph)}. The execution will terminate whenever \emph{boolean continueGeneration()} returns false.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.55\linewidth]{validationresultlistener_class} 
  \caption{Interface definition of \emph{ValidationResultListener}.}
  \label{fig:validationresultlistener_class}
\end{figure}

@@

Figure~\ref{fig:editor} shows a snapshot of the editor's environment: All functionality needed for constraint editing is provided in the tool bar on the right. It contains draggable elements for creating all operator and proposition types as well as a trashcan for deleting. Constraints can be composed by drag\&drop in the dashboard in the center of the editor which holds and displays the operator constraints. The tab functionality on the left allows multiple constraints to be managed. Each tab shows a small thumbnail of the constraint and a symbol indicating its validity; a warning shield for ``incomplete'', a green shield for ``valid'', a red cross for ``invalid'' or an animated ring for ``validation in progress.''. The magic wand button within the tab pane makes the constraint generation functionality accessible within the editor and triggers the constraint finding process.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\linewidth]{editor} 
  \caption{Snapshot of the visual editor.}
  \label{fig:editor}
\end{figure}

@@

LTLSPEC G (state = 5 -> (F (state = 8 | state = 11)))
\end{lstlisting}
A state machine is encoded in one variable \emph{state}, and its value represents the active state. Thus each state is identified by a number. The state machine described in the example contains eleven states what is determined by the number range for \emph{state} from one to eleven. All variable assignments are specified under \emph{ASSIGN} and represent the semantic equivalent to transition definitions. The initial state can be determined and a case distinction regulates the next possible values, i.e. transitions to reachable states. If there is more than one outgoing transition, all target state identifiers have to be surrounded by brackets. The last expression \emph{TRUE : state;} is a self transition for every state not considered in the case distinction.
The last line contains the constraint formula we want to be checked on the state machine. The prefix \emph{LTLSPEC} just says that NuSMV has to interprete the following constraint as LTL formula.

@@

It can happen that there is more than one validation task at a time. This is the case when the constraint generator creates multiple constraints or a change of the state machine causes all constraints to be revalidated at once. Model checking is a quite ressource inefficient activity and parallel execution of several checks might overload the main memory what causes slow data swapping to start. Since each constraint is checked in a new instance of NuSMV, a parallel execution would furthermore strain the CPU scheduler and slow down the overall execution time. Thus it's recommended to execute all validation tasks consecutively rather than in parallel.
An extra advantage is that the developer doesn't have to wait for all tasks to finish before he can see any result. Since a validation task is not startet until the previous one is finished, constantly new validation results can be displayed to the developer.
All this is managed by the \emph{ValidatorThread}, a queue based mechanism for scheduling validations. Its class definition is presented in figure~\ref{fig:validatorthread_class}.
\begin{figure}[htbp]
  \centering
  \includegraphics[width=\linewidth]{validatorthread_class} 
  \caption{Class definition of \emph{ValidatorThread}.}
  \label{fig:validatorthread_class}
\end{figure}
Every validation task of a constraint gets enqueued in the \emph{ValidatorThread} after calling the method \emph{addToQueue(...)}.
An identifier is used for each constraint which enables elimination of duplicate tasks for the same constraint in order not to strain the execution time. Tasks can also manually be removed from the queue by calling \emph{cancel(...)}. After a validation task is finished the result is published via callback to the \emph{ValidationResultListener}.

@@

LTLCreator should be integratable into other state machine based programming environments, what is ensured by its implementation as a Java Swing component and a clear API enabling a wide range of use of the tool.

>>>>

First of all the model implementation for representing state machines is described in section~\ref{sec:statemachinemodel}. Then, some details about the implementation of the visual language are given in section~\ref{sec:prototype:operatorconstraints}, directly followed by an explanation about the realization of the automated constraint generation functionality in section~\ref{sec:prototype:automatedconstraintgeneration}.
The assembly of all the previous features in one tool is described in section~\ref{sec:coalescence}. However, the tool's functionality shall also be available in Robostudio in order to provide safety mechanisms for the healthcare service robots and to be a prototype for evaluation. Section~\ref{sec:integrationintorobostudio} gives a main idea how an integration into Robostudio is possible.

\section{State machine model}
\label{sec:statemachinemodel}

For the checking of constraints as well as for the constraint generation a model is needed which represents the state machine program. Three classes are responsible for holding the state machine program with all its necessary information: The class \emph{Fsm} represents the finite state machine and lists all contained states. It is called finite state machine, because the number of states is finite which applies to the domain of this work. \emph{State} represents the states of the machine and holds all transitions from and to the respective state. Transitions are represented by the class \emph{Transition} which specifies the source and a target state. Figure~\ref{fig:fsm_class} shows all three class definitions for the state machine model.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.6\linewidth]{fsm_class} 
  \caption{State machine model class definitions.}
  \label{fig:fsm_class}
\end{figure}

@@

\begin{description}
	\item[getLTL():] This function returns the LTL formula of an operator and all its children recursively. Applied to the most outer operator it returns the complete formula for the constraint. This function is abstract and implemented by each operator type separately.
	\item[getColor():] Each operator type has got is own significant color which can be retrieved by the abstract function \emph{getColor()}.
	\item[createNewInstance():] In order to provide operator creation over a tool bar, this function instantiates a new operator object, depending on the implementation.
	%\item[setMouseOver(boolean):] Mouse over actions cause the underlying operator to be highlighted. This function makes all operators except the hovered one appearing bleeched out.
	%\item[paintOperator(Graphics, ...):] Each operator class has to implement the abstract function \emph{paintOperator()}. It manages all color and shape painting for the respective operator. Additional parameters such as recursive painting including children can be specified.
	%\item[contains(int, int):] Given two coordinates x and y, it can be determined whether the herewith defined point lays within an operator.
	\item[isSimilar(AbstractOperator):] Two operators can be compared for semantic equality. Equality is given when the compared operators are of the same type and similarly their children. This functionality is important for avoiding duplicate constraints during constraint generation, for example.
	\item[addChangeListener(OperatorChangeListener):] Whenever there is a change to the operator or one of its children, an operator change listener gets notified about it. The hierarchical architecture of \emph{AbstractOperator}s propagates all change events to the root operator. The function \emph{addChangeListener()} allows registering for such change notifications.
	\item[removeChangeListener(OperatorChangeListener):] According to \emph{addChangeListener} this function allows deregistration again.
\end{description}

@@

Whereas \emph{StateOperator}s have no operator children and are leafs in the syntax tree of LTL formulas, all other operator types are either unary or binary operators with one or two possible nested children respectively. 
The graphical representation of both unary and binary operators thus needs to have place holders and docking stations
\begin{wrapfigure}{r}{0.40\textwidth}
  \centering
  \includegraphics[scale=0.65]{bucket_example.jpg}
  \caption{Buckets for graphical nesting of operators.}
  \label{fig:bucket_example}
\end{wrapfigure}
for sub operators. These are realized by buckets which display a small ``drop'' label as shown in figure~\ref{fig:bucket_example} and allow operator adding or removing by mouse drops.
For this a class \emph{Bucket} is defined with get and set methods for defining a sub operator.
Figure~\ref{fig:klassendiagramm_operators} depicts the relationship between \emph{AbstractOperator}, \emph{Bucket} and operator implementations.
Once operators are composed in a way that no empty buckets are left over, a complete constraint such as the one shown in figure~\ref{fig:operator_tree} is retrieved. This diagram illustrates the correlation of nested operators and buckets.
\begin{figure}[htbp]
  \centering
  \includegraphics[width=\linewidth]{klassendiagramm_operators} 
  \caption{Class diagram of operator structure.}
  \label{fig:klassendiagramm_operators}
\end{figure}

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.95\linewidth]{operator_tree} 
  \caption{Instantiation of a constraint with several operators.}
  \label{fig:operator_tree}
\end{figure}

For the prototype implementation, all graphical components such as buckets and operators are implemented as Java Swing \emph{JComponent}. Thereby a lot of functionality for the visual configuration is already available such as the complex layout and repaint concept and the container functionality which is used for unary and binary operators as well as buckets.

@@

Swing allows the implementation of a \emph{paint()} method which is responsible for rendering the component. This is used for giving every operator its special look. Swing's \emph{Graphics2D} functionality provides a great support in painting rounded shapes and color gradients. An operator is displayed as a rounded rectangle with dents around sub operators and has a flowing backgound, borders and labels in an operator type specific color. For the significant shape multiple rounded rectangles with and without borders are arranged in a way that the result appears to be one piece. Figure~\ref{fig:paintprocess} demonstrates how the process of painting is: For painting the \emph{IfThenOperator}, first of all a blank rectangle with border is drawn, directly followed by two filled rectangles with border which are placed where later the two child operators shall be. In the third step the same rectangle as from step one is painted on top, this time without border. This makes all crossing borders disappear. Only labels and buckets still have to be added to create a complete \emph{IfThenOperator}.

@@

As proposed in the requirements for the visual language in section~\ref{sec:operatorconstraintformalism}, constraints have particular meanings for the two dimensions: A vertical read direction and the horizontal direction for variation in time. 
Swing's mechanism for layouting components is used for the positioning of operators. After determining the aspired sizes of child components which represent sub operators, reference lines which have been already demonstrated with figure~\ref{fig:referencelines} in section~\ref{sec:operatorconstraintformalism} are computed mathematically. Based on this information operators are sized and positioned correctly.

%MOUSE OVER

Regarding to the requirement for simple reading of constraints based on mouse hovering, the editor prototype supports context sensitive highlighting of particular constraint parts.
The built-in mouse support of Swing is used for identifying the operator positioned under the mouse cursor. All operators except the hovered one and its sub operators are presented bleached out what causes the hovered operator to appear emphasized. The most right operator composition in figure~\ref{fig:nested_and_mouseover} gives a demonstration of this idea.

%PROBLEM WITH MOUSE OVER AND MERGING

However, the concept of operator highlighting causes trouble in combination with the graphical merging of visual operators which was introduced in section~\ref{sec:designdecisions}. Since a merged operator does not have its own border, it can not be displayed as desired when hovered. This fact is also a problem for the editing: The graphical presentation does not support changing of merged operators, because it seems that they can not be picked by the mouse.
In order to retain full editability, hovered operators are excluded from the merge rule and still have to be displayed separated (see~figure~\ref{fig:nested_and_mouseover}).
\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.65]{and_simplify_mouseover}
  \caption{Default nested \emph{AND} operator, visually merged nested \emph{AND} operator and visually nested \emph{AND} operator with mouse over.
  }
  \label{fig:nested_and_mouseover}
\end{figure}

@@

Besides its importance for mouse hovering, Swing's mouse support is also used for determining the position where operators have to be added after drag and drop edit actions. So far
% the predefined Swing drag and drop has not been used but
a simple drag and drop mechanism called \emph{DndHandler} has been implemented whose class structure is shown in figure~\ref{fig:dndclassstructure}.
\begin{figure}[htbp]
  \centering
  \includegraphics[width=\linewidth]{dndclassstructure} 
  \caption{Class definition for drag and drop functionality.}
  \label{fig:dndclassstructure}
\end{figure}% instead. 
Once a \emph{JComponent} is registered to the \emph{DndHandler}, it automatically manages any drag and drop actions with \emph{JComponent}s which implement either the \emph{Draggable} or \emph{DropTarget} interface. The former is implemented by all operators and specifies that this object can be moved from and to buckets or the dashboard, the latter is implemented by buckets and the dashboard and determines that it can accept and contain \emph{Draggable}s such as operators.
%Figure~\ref{fig:dndclassstructure} gives a short class overview.
Furthermore there is a class \emph{DraggableProvider} which is similar to \emph{Draggable} except that it is not moved itself but creates a new instance of the respective \emph{Draggable} implementation type which is then moved instead.

%TOOLBAR

\emph{LLTCreator} is equipped with a tool bar which provides all functionality needed for constraint editing: For each operator type a \emph{DraggableProvider} is displayed as an icon in the respective operator color. Each time the mouse is pressed on it (drag action start), a new instance of the respective operator type is created which can be moved and dropped to the dashboard. Additionally, a trash can icon is displayed within the tool bar. It implements the \emph{DropTarget} interface and can thus receive operators. Its purpose is to provide a remove functionality for already instantiated operators which are not needed anymore.

@@

As stated in section~\ref{sec:subgraphdefinition}, the automated generation of constraints is based on subgraphs. For the finding of such subgraphs an algorithm has been developed. Section~\ref{sec:subgraphfindingconcept} aims for explaining the basic concept of the algorithm, section~\ref{sec:subgraphfindingimplementation} introduces the slightly modified algorithm which is used for the prototype.
%It is described as follows:

\subsection{Subgraph finding concept}
\label{sec:subgraphfindingconcept}

First of all one start state ($\alpha$) is defined which has two or more outgoing transitions and all possible paths are generated which start from it and lead through the state machine. Excluded are all paths which contain states more than once.
The result is a finite amount of finite paths since loops are eliminated. Now every state gets assigned a number which represents the total number of occurrences of the respective state in all collected paths. In other words, the label of a state expresses the number of paths which start in the $\alpha$-state and lead through the respective state. The $\alpha$-state itself is labeled with the total number of paths. If during walking through any of the paths a second state can be found with the same label, it means that all paths starting at $\alpha$-state come all together in this state again. It is determined to be a potential corresponding end state ($\beta$).
A last check has to be applied whether there are incoming transitions from outside into the subgraph which do not originally come from $\alpha$-state. If no such transitions are detected, a subgraph is found.
If there is no state with the same label, or transitions come from outside, the state machine as a whole can be assumed as subgraph beginning with $\alpha$-state insofar as the state machine does not contain further loops or terminating end states. Figure~\ref{fig:subgraph_finding} illustrates this concept of subgraph finding and shows all possible paths as well as the number of occurrence of each state.
\begin{figure}[htbp]
  \centering

@@

	\end{tikzpicture}
  \caption{Example for subgraph finding. The occurrence of each state in the paths lets you determine G as an end state ($\beta$) corresponding to A.}
  \label{fig:subgraph_finding}
\end{figure}
This procedure is suitable for checking whether a state is the beginning of a subgraph and for discovering the corresponding $\beta$-state. In order to detect all subgraphs within a state machine, this procedure has to be applied to every single state wich has two or more outgoing transitions. In order to improve the further finding of subgraphs and walking through the state machine, all already found subgraphs can be assumed as single states since it is already known where their paths come finally together again: In their $\beta$-states.

\subsection{Subgraph finding implementation}
\label{sec:subgraphfindingimplementation}

Following the previous description an algorithm has been implemented in Java. Its realisation differs from the presented concept in one point: Paths are not listed but states are directly labelled since not saving all possible paths is more memory efficient.
The state machine is traversed similarly to a depth-first search and each visited state's label is incremented by one. If the currently visited state holds more than one outgoing transition and thus constitutes new paths,  all states on the current stack, i.e. the previously visited states on the current path, get their labels incremented by $(\textnormal{'number of outgoing transitions'} - 1)$. The algorithm turns back whenever it comes across a state which has already been visited within the same path stack of the depth-first search.

The overall implemented way of subgraph finding is illustrated in pseudo code with algorithm~\ref{alg:subgraphfinding}.
\begin{algorithm}
\caption{Finding of all subgraphs.}
\label{alg:subgraphfinding}
\begin{algorithmic}[1]
\STATE $subgraphs$: List
\FORALL{states as $S$}
  \STATE $start \leftarrow S$
	\STATE doLabelling($start$);
	\IF{state $E$ exists with label($E$)$=$label($S$)}
		\STATE $end \leftarrow E$
	\ELSE
		\STATE $end \leftarrow start$
	\ENDIF
	\IF{noCycles($start$, $end$) $and$ noExtTransitions($start$, $end$)}
		\STATE $subgraphs$.add($start$, $end$);
	\ENDIF
\ENDFOR
\end{algorithmic}
\end{algorithm}
In line 4 the labelling process described above is executed by using one particular start state. If a state has the same label as the start state, it is assumed as potential end state. Otherwise, the entire state machine is assumed to be a subgraph and start state is defined as end state as well. Subsequent checks finally determine whether it can be considered as proper subgraph.
There are no cycles allowed between start and end state since they are forbidden by the subgraph definition in section~\ref{sec:subgraphdefinition}.
Furthermore also foreign transitions are prohibited. Transitions coming from a state outside the subgraph are allowed to lead to the start state but not to other states of the subgraph. If all checks are successful, start and end states get added to the result list $subgraphs$. This procedure is applied to all states of the state machine by a surrounding foreach loop. In the end, the list $subgraphs$ contains all pairs of start and end states which form subgraphs.

\subsection{Constraint generation}

After determining a subgraph, constraints are derived from it as defined in section~\ref{sec:constraintdetermination}. The entire process of constraint generation is illustrated by figure~\ref{fig:processofconstraintgeneration}. Once the user triggers the constraint generation, the subgraph finding algorithm starts working. When a subgraph is found, constraints are created and passed to the user. For each subgraph multiple constraint types can apply. This procedure is done iterative for each subgraph found.

\begin{figure}
  \centering

  \begin{sequencediagram}
		\tikzstyle{inststyle}+=[top color=yellow!30,bottom
color=white]
		\newthread{u}{User/Editor}{blue}
		\newthread[2.6]{cc}{Constraint creator}{red}
		\newinst[1]{alg}{Subgraph finding algorithm}

		\mess{u}{start(stm)}{cc}
		\begin{call}[3]{cc}{start(stm)}{alg}{finish}
			\begin{sdblock}[yellow!20]{For each subgraph}{}
				\setthreadbias{east}
				\begin{call}[3]{alg}{newSubgraph(subgraph)}{cc}{}
					\begin{sdblock}[yellow!20]{For each constraint}{}
						\begin{call}[3]{cc}{newConstraint(constraint)}{u}{}
						\end{call}
					\end{sdblock}
				\end{call}
				\setthreadbias{center}
			\end{sdblock}
		\end{call}
		\mess{cc}{finish}{u}
	\end{sequencediagram}

  \caption{Process of constraint generation.}
  \label{fig:processofconstraintgeneration}
\end{figure}

The runtime of the algorithm depends on size and grade of branch of the state machine. An estimation about the execution time is presented later in section~\ref{sec:performance}. In order to improve waiting times and thus user experience during constraint generation the entire result might not be returned after termination but each subgraph should be processed as soon as it is found.
For this, the algorithm implementation has been extended by code snippets which frequently interact with so called \emph{ResultListener}s (figure~\ref{fig:resultlistener_class}).
%This interface is also implemented by the progress dialog.
An implementation of \emph{ResultListener} has to provide three methods which allow the algorithm to publish the current progress by calling \emph{void setProgress(int)} or to notify about newly found subgraphs by invoking \emph{void newSubGraphFound(SubGraph)}. The method \emph{boolean continueGeneration()} is called frequently and the execution will terminate as soon as it returns false.
\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.55\linewidth]{resultlistener_class} 
  \caption{Interface definition of \emph{ResultListener}.}
  \label{fig:resultlistener_class}
\end{figure}

As a second approach aiming for better user experience during constraint generation, a dialog has been implemented which informs the developer continuously about the current progress. This includes the number of constraints found so far and a progress bar which gives a percentual estimation of execution time. Furthermore a cancel button permits premature stopping of the execution. A snapshot of the Dialog is shown in figure~\ref{fig:progressdialog}.
\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.40\linewidth]{progressdialog} 
  \caption{Snapshot of the progress dialog displayed during constraint generation.}
  \label{fig:progressdialog}
\end{figure}

@@

Figure~\ref{fig:editor} shows a snapshot of the editor's environment:
\begin{figure}[htbp]
  \centering
  \includegraphics[width=\linewidth]{editor} 
  \caption{Snapshot of the visual editor.}
  \label{fig:editor}
\end{figure}
All functionality needed for constraint editing is provided in the tool bar on the right. It contains draggable elements for creating all operator and proposition types as well as a trashcan for deleting. Constraints can be composed by drag\&drop in the dashboard in the center of the editor which holds and displays the operator constraints. The tab functionality on the left allows multiple constraints to be managed. Each tab shows a small thumbnail of the constraint and a symbol indicating its current status of validation result: For syntactically invalid constraints, an ``incomplete'' sign is displayed. Otherwise, an animated ring indicates that validation is in pro\-gress and will finally result in either a ``valid'' or ``invalid'' sign as shown in figure~\ref{fig:results}.
\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.65]{results}
  \caption{Status indicators displaying validation results: incomplete, validation in pro\-gress, invalid and valid.}
  \label{fig:results}
\end{figure}	
The magic wand button within the tab pane makes the constraint generation functionality accessible within the editor and triggers the constraint finding process.

@@

LTLSPEC G (state = 5 -> (F (state = 8 | state = 11)))
\end{lstlisting}
A state machine is encoded in one variable \emph{state}, and its value represents the active state. Thus, each state is identified by a number. The state machine described in the example contains eleven states, this is determined by the number range for \emph{state} from one to eleven. All variable assignments are specified under \emph{ASSIGN} and represent the semantic equivalent to transition definitions. The initial state can be determined and a case distinction regulates the next possible values, i.e. transitions to reachable states. If there is more than one outgoing transition, all target state identifiers have to be surrounded by brackets. The last expression \emph{TRUE : state;} is a self transition for every state not considered in the case distinction.
The last line contains the constraint formula we want to be checked on the state machine. The prefix \emph{LTLSPEC} just says that NuSMV has to interprete the following constraint as LTL formula.

@@

It can happen that there is more than one validation task at a time. This is the case when the constraint generator creates multiple constraints or a change of the state machine causes all constraints to be revalidated at once. Model checking is a quite ressource inefficient activity and parallel execution of several checks might overload the main memory what causes slow data swapping to start. Since each constraint is checked in a new instance of NuSMV, a parallel execution would furthermore strain the CPU scheduler and slow down the overall execution time. Thus it is recommended to execute all validation tasks consecutively rather than in parallel.
An extra advantage is that the developer does not have to wait for all tasks to finish before he can see any result. Since a validation task is not startet until the previous one is finished, constantly new validation results can be displayed to the developer.
All this is managed by the \emph{ValidatorThread}, a queue based mechanism for scheduling validations. Its class definition is presented in figure~\ref{fig:validatorthread_class}.
\begin{figure}[htbp]
  \centering
  \includegraphics[width=\linewidth]{validatorthread_class} 
  \caption{Class definition of \emph{ValidatorThread}.}
  \label{fig:validatorthread_class}
\end{figure}
Every validation task of a constraint gets enqueued in the \emph{ValidatorThread} after calling the method \emph{addToQueue(...)}.
An identifier is used for each constraint which enables elimination of duplicate tasks for the same constraint in order not to strain the execution time. Tasks can also manually be removed from the queue by calling \emph{cancel(...)}. After a validation task is finished the result is published via callback to the \emph{ValidationResultListener}.

@@

LTLCreator should be integratable into other state machine based programming environments. This is ensured by its implementation as a Java Swing component and a clear API which enables a wide range of use for the tool.
